`respR` - An R package for the analysis of respirometry data

### 2017-10-19 manuscript/manuscript-HTML.Rmd
manuscript update - with edits included

There are four broad methodological approaches to respirometry: closed, intermittent flow, flowthrough and open. In closed respirometry, $O_2$ decrease is measured within a hermetically sealed chamber of known volume, sometimes set within a closed loop to allow mixing of the environment within the chamber. Oxygen recordings may be continuous through use of an oxygen probe, periodic through withdrawing water or gas samples at set intervals, or a two-point measurement consisting of the initial and final concentrations. Metabolic rates are estimated from respirometry data by assuming a linear relationship between variables, and estimates of metabolic rate are straightforward in constant volume respirometry using the equation: $$VO_2 = \Delta O_2V$$ where $\Delta O_2V$ is the slope of the regression that describes the rate of change in $O_2$ concentration over time, or in the case of a two-point measurement, the difference in $O_2$ concentration divided by time elapsed, and $V$ is the volume of the container (Lighton, 2008). ~~Note, we use here and in the package the term $RO_2$ to denote rate of $O_2$ depletion or uptake within an experiment. We use this term used to distinguish it from other common terms, for example $VO_2$ which is can be used to denote both volumetric O2 uptake and whole animal uptake, or $MO_2$ which typically denotes mass-specific uptake~~ [nc - don't need this anymore. I had it when we used that term in my functions].

In intermittent flow respirometry, $O_2$ concentration is measured as described above, but periodically the chamber is flushed with new water or air, returning it to initial conditions, resealed, and the experiment repeated (Svendsen et al., 2016). This technique is essentially closed respirometry, but with the incorporation of repeated measures. Depending on the metabolic rate metric being investigated, final respiration rate can be calculated as the mean of the measures (e.g. Carey et al., 2016), or the lowest or highest rates recorded in any trial (e.g. Stoffels, 2015). [nc-will try and find a better ref for that]

Stoffels, R. J. (2015). Physiological trade-offs along a fast-slow lifestyle continuum in fishes: What do they tell us about resistance and resilience to hypoxia? Plos One, 10(6), e0130303. https://doi.org/10.1371/journal.pone.0130303

@@

[nc - so the following needs to be spot on. Physiologists get very worked up (if not confused themselves) about the differences between these, esp. SMR v Routine v Resting v Basal! Especially those who work on ecto vs endo therms]

Depending on the experiment, different metabolic rate metrics may be of interest to researchers. These include maximum metabolic rates (MMR, or $MO_{2,max}$), rates under high activity or exhaustive exercise, and minimal metabolic rates $MO_{2,min}$, which depending on the organism may be termed standard (SMR), basal (BMR), or resting metabolic rates, and typically represent the minimum metabolic cost of maintaining biological functioning (White and Kearney, 2014) [much better ref for all this]. Routine metabolic rate (RMR), is similar but takes into account that in some organisms energy is expended on small, sponateous movements to maintain posture or position (Rogers et al. 2016). Another metric of great interest is the critical oxygen tension, which represents the lowest level of oxygen at which aerobic metabolism is independent of the ambient partial pressure of oxygen ($P_{crit}$; Yeager & Ultsch, 1989, Hochachka and Somero, 2002).

White, C. R., & Kearney, M. R. (2014). Metabolic scaling in animals: methods, empirical results, and theoretical explanations. Comprehensive Physiology, 4(1), 231–256. https://doi.org/10.1002/cphy.c110049

Researchers are increasingly assimilating large, high-resolution respirometry datasets. Experiments may run for long periods (e.g. 20 h, Norin & Malte, 2012), and in most cases, processing the data involves an *ad hoc* selection of data points under vague criteria, with subsequent manual processing in a spreadsheet program (e.g. Microsoft Excel) or an integrated development environment (IDE, e.g. R and Matlab). These approaches can be tedious and time consuming especially when spreadsheet programs (in particular Excel) struggle with the vast datasets that are generated, while IDEs require a degree of expertise and have substantial learning curves. Dedicated software are available to perform such analyses, but many have expensive licensing requirements (e.g. AutoResp by Loligo Systems), complicating or preventing their use on multiple machines, and are proprietary and closed-source hindering scientific reproducability and transparency.

[r.e. below para. LoLinR is not strictly for respirometry - just monotonic, something-changing-over-time data]

Recently, a number of open-source R software packages have become available (e.g. `respirometry`: https://CRAN.R-project.org/package=respirometry, `rMR`: https://CRAN.R-project.org/package=rMR, `LoLinR`: Olito et al., 2017) which are suitable for or designed to analyse respirometry data. ~~Both `respirometry` and `rMR` are useful for aquatic respirometry, but are not focused on the analysis of data, although both packages have functions for the analysis of metabolic rate over non-overlapping intervals, and `rMR` has a specialised function to calculate $P_{c}$.~~ [nc - need a better way of basically saying, they're ok, but very limited, and not user-friendly, or adaptable, or]. `LoLinR` provides a very statistically robust method of detecting a "best-fit" regression which is applicable for closed and flow-through respirometry. [nc - try to add something else positive about it! We need to be nice to them]. However, `LoLinR`'s processing time increases exponentially with large datasets, with data containing more than 500 observations taking several minutes to run (Olito et al., 2017), which makes `LoLinR` somewhat intractable with large amounts of data. In addition, its focus is on data which is evenly spaced with regard to time, and where the response variable changes in a monotonic fashion, which is far from typical of respirometry data.

Here we describe the R package `respR`, a set of functions desgined to provide a fast [or better word], structured, reproduceable workflow for the analysis of respirometry-related data (Fig. 1). The package contains utilities to: (1) analyse closed, intermittent, flow-through and open respirometry data, (2) determine volume-, mass-, and surface area- [nc - soon yes?] specific oxygen uptake rates, (3) automatically detect maximal and minimal rates, $P_{c}$, or the most linear results in large datasets with the help of traditional rolling regression and kernel density estimation techniques. Other functions useful in processing respirometry data are avaliable, and are described in more detail in our online html vignette (https://januarharianto.github.io/respR/). The package has a strong focus on aquatic respirometry, however similar principals apply to aerial respirometry, so the analysis workflow should also be suitable for these data. ~~Some oxygen concentration conversions require a salinity value, but in freshwater respirometry this should be set to zero.~~ [nc - don't need this, can go in the help docs]. We demonstrate the utility of the package by analysing three datasets as case studies and proof of concept. These data were collected using closed, intermittent and flow-through respirometry.

[nc - i would drop closed, focus on intermittent (which is same as closed essentially) and flowthrough, and include a PCrit case. It's a major selling point! Several people i have told about it have asked - does it do PCrit? Or i guess, make the closed e.g. the PCrit one... ]

@@

Data should be formatted correctly before use in `respR`. The function `check.input()` performs error checks on a data frame of any size, and can extract a two-column data frame for data exploration and subsequent analyses. Data must be numeric, because `respR` has limited support for date/time data (e.g. `POSIXct` and `POSIXlt` classes) and performing calculations on different time formats have many caveats that are beyond the scope of the package. Time data, being a continuous variable, should be sequential, without duplicates. Ideally, time data should also be evenly-spaced, that is, the time period between samples are equal, as this has important ramifications on data extraction methods later on, and substantial errors may occur if data is analysed by rows instead of time. The function summarises error checks in the R console (see supplementary material) and plot of the data is shown for visual inspection (e.g. Fig. 1).

@@

## Estimating $RO_2$

The functions `calc.bg.rate()`, `calc.rate()`, `auto.rate()`, `calc.rate.ft()` and `pcrit()` are used to process raw respirometry data. For delibrate reasons, units of measurements are not required in any of these functions.

### Analysing background changes in $RO_2$

@@

### Manual selection of data

Typically, the analysis respirometry data requires the selection one or more linear sections of the raw data for more accurate calculations of $RO_2$ (e.g. Gordon et al., 1989, Dorey et al., 2014, Chu & Gal, 2017). The function `calc.rate()` can manually extract and process data segments from a `data.frame` object. Data segments can be selected by (1) time period, (2) row numbers, (3) $O_2$ decrease or (4) proportion, which should accomodate most, if not all, data selection requirements and allow for consistent reporting of methods and results (e.g. estimating $RO_2$ over a regular 10-minute time window across multiple specimens). Rates are determined using linear regression, however a simple two-point analysis (equation 1) is also included alongside all regression results.

### Automatic selection of data

The function `auto.rate()` uses rolling regression and kernel density estimate algorithms to detect patterns in respirometry data. It can calculate maximum or minimum rates, extract rates at non-overlapping intervals, and has an exploratory method to detect the most linear sections of the data.

There are clear advantages to using `auto.rate()` to estimate MMR and SMR. Unlike manual methods, where data is often arbitrary selected and inconvenient to replicate, `auto.rate()` adheres to well-defined rules to obtain the rate of interest, so its methods are results are transparent and fully reproducible. When called, the function takes the data frame of length $m$, and performs a rolling regression of sample size $n$ iteratively across the length of the data (Fig. 2b). Thus, a total of $(m−n)+1$ number of overlapping regressions are fitted, and we obtain a new dataset of regressions as a function of time (rolling regression, Fig. 3b), which is then ranked by size to obtain maximum and minimum values.

While using rolling regressions to estimate MMR and SMR

### Analysing flowthrough respirometry data

The function `calc.rate.ft()` is used to analyse flowthrough respirometry data, and accepts inputs of incurrent and excurrent oxygen concentration, and flow rate. It is crucial that all the input units match as the function does not perform unit conversions (but see `calc.mo2()`), and the output gives $RO_2$. Like `calc.rate()`, a segment of the data can be selected before processing and the average, or median $VO_2$ are provided.

@@

Once

### Critical oxygen tension, $P_{c}$

- oxygen tolerance (Hochachka and Somero, 2002)
- broken stick model (Yeager and Ultsch, 1989)
- $RO_2$ vs $P_{O_2}$ data can be generated automatically from raw $O_2$ vs time data.

The function `pcrit()` provides a way to consistently report the parameters associated with estimating $P_c$. $P_c$ is often estimated using the "broken-stick" regression (BSR) approach (Yeager and Ultsch, 1989), although there are some cases where it is estimated without indication of how thie value was determined numerically (e.g. need examples). While in many cases the PCrit and oxyregulatory ability can be clear and obvious (e.g./refs), many studies still class species or individuals as oxyregulators or oxyconformers without recognising that there may be intermediate cases.

>>>>

There are four broad methodological approaches to respirometry: closed, intermittent flow, flowthrough and open. In closed respirometry, $O_2$ decrease is measured within a hermetically sealed chamber of known volume, sometimes set within a closed loop to allow mixing of the environment within the chamber. Oxygen recordings may be continuous through use of an oxygen probe, periodic through withdrawing water or gas samples at set intervals, or a two-point measurement consisting of the initial and final concentrations. Metabolic rates are estimated from respirometry data by assuming a linear relationship between variables, and estimates of metabolic rate are straightforward in constant volume respirometry using the equation: $$VO_2 = \Delta O_2V$$ where $\Delta O_2V$ is the slope of the regression that describes the rate of change in $O_2$ concentration over time, or in the case of a two-point measurement, the difference in $O_2$ concentration divided by time elapsed, and $V$ is the volume of the container (Lighton, 2008). 
<!-- Note, we use here and in the package the term $RO_2$ to denote rate of $O_2$ depletion or uptake within an experiment. We use this term used to distinguish it from other common terms, for example $VO_2$ which is can be used to denote both volumetric O2 uptake and whole animal uptake, or $MO_2$ which typically denotes mass-specific uptake. -->

In intermittent flow respirometry, $O_2$ concentration is measured as described above, but periodically the chamber is flushed with new water or air, returning it to initial conditions, resealed, and the experiment repeated (Svendsen et al., 2016). This technique is essentially closed respirometry, but with the incorporation of repeated measures. Depending on the metabolic rate metric being investigated, final respiration rate can be calculated as the mean of the measures (e.g. Carey et al., 2016), or the lowest or highest rates recorded in any trial (e.g. Stoffels, 2015). 
<!--[nc-will try and find a better ref for that]

Stoffels, R. J. (2015). Physiological trade-offs along a fast-slow lifestyle continuum in fishes: What do they tell us about resistance and resilience to hypoxia? Plos One, 10(6), e0130303. https://doi.org/10.1371/journal.pone.0130303
-->

@@

<!--[nc - so the following needs to be spot on. Physiologists get very worked up (if not confused themselves) about the differences between these, esp. SMR v Routine v Resting v Basal! Especially those who work on ecto vs endo therms]-->

Depending on the experiment, different metabolic rate parameters may be of interest to researchers. These include maximum metabolic rates (MMR, or $MO_{2,max}$), rates under high activity or exhaustive exercise, and minimal metabolic rates ($MO_{2,min}$), which, depending on the organism, may be termed standard (SMR), basal (BMR), or resting metabolic rates, and typically represent the minimum metabolic cost of maintaining biological functioning (White and Kearney, 2014). Routine metabolic rate (RMR), is similar to $MO_{2,min}$, but takes into account that in some organisms energy is expended on small, sponateous movements to maintain posture or position (Rogers et al. 2016). Another metric of great interest is the critical oxygen tension, which represents the lowest level of oxygen at which aerobic metabolism is independent of the ambient partial pressure of oxygen ($P_{crit}$; Yeager & Ultsch, 1989, Hochachka and Somero, 2002).

<!--White, C. R., & Kearney, M. R. (2014). Metabolic scaling in animals: methods, empirical results, and theoretical explanations. Comprehensive Physiology, 4(1), 231–256. https://doi.org/10.1002/cphy.c110049-->

Respirometry experiments are increasingly assimilating large, high-resolution datasets and are run for longer periods (e.g. 20 h, Norin & Malte, 2012). In most cases, processing the data involves an *ad hoc* selection of data points with undefined criteria, and subsequent manual processing of the data subset(s) using a spreadsheet program (e.g. Microsoft Excel) or an integrated development environment (IDE, e.g. R or Matlab). These approaches can be tedious and time consuming especially when spreadsheet programs struggle with the vast datasets that are generated, while IDEs require a degree of expertise to use and have substantial learning curves. Dedicated software are also available to perform metabolic rate analyses, but many have costly licensing requirements (e.g. AutoResp by Loligo Systems), complicating or preventing their use on multiple machines, and are proprietary and closed-source, hindering scientific reproducability and transparency.

<!-- r.e. below para. LoLinR is not strictly for respirometry - just monotonic, something-changing-over-time data -->

Recently, a number of open-source R software packages have become available which are designed for, or are suitable to analyse respirometry data. The `respirometry` package (https://CRAN.R-project.org/package=respirometry) contains a comprehensive collection of tools to explore and evaluate experimental parameters in aquatic respirometry and is useful for the design and diagnosis of experimental setups. The `rMR` package (https://CRAN.R-project.org/package=rMR) performs interval-based metabolic rate calculations and has a method to automatically detect $P_{crit}$ using the "broken-stick" regression method (Yeager & Ultsch, 1989). `LoLinR` (Olito et al., 2017), while not strictly coded with respirometry in mind, provides a statistically robust method of detecting a "best-fit" regression that applies to the calculation of metabolic rate, and performs very well at identifying truly linear subsets of a data series. Importantly, the packages above all assume that data are monotonic with evenly-spaced sampling periods with respect to time and do not have the tools to analyse complicated datasets that require extraction of multiple respirometry parameters.

Here we describe the R package `respR`, a set of functions desgined to provide an efficient, structured, and reproduceable workflow for the analysis of respirometry-related data (Fig. 1). The package contains utilities to: (1) analyse closed, intermittent, flow-through and open respirometry data, (2) determine volume, mass, and surface area specific oxygen uptake rates, (3) automatically detect maximal and minimal rates, $P_{crit}$, or the most linear results in large datasets with the help of traditional rolling regression and kernel density estimation techniques. Other functions useful in processing respirometry data are avaliable, and are described in more detail in our online html vignette (https://januarharianto.github.io/respR/). The package has a strong focus on aquatic respirometry, however similar principals apply to aerial respirometry, so the analysis workflow should also be suitable for these data. We demonstrate the utility of the package by analysing three datasets as case studies and proof of concept. These data were collected using closed, intermittent and flow-through respirometry, and we show how `respR` might be used to determine $MO_{2,max}$, $MO_{2,min}$, and $P_{crit}$.

@@

Data should be formatted correctly before use in `respR`. The function `check.input()` performs error checks on a data frame of any size, and can extract a two-column data frame for data exploration and subsequent analyses. Data must be numeric, because `respR` has limited support for date/time data (e.g. `POSIXct` and `POSIXlt` classes) and performing calculations on different time formats have many caveats that are beyond the scope of the package. Time data, being a continuous variable, should be sequential, without duplicates. Ideally, time data should also be evenly-spaced, that is, the time period between samples are equal, as this will affect data-extraction methods later on if data is analysed by sample length instead of time. Error check results are summarised in the R console (see supplementary material) and plot of the data is shown for visual inspection (e.g. Fig. 1).

@@

## Estimating the rate of change in $O_2$ concentration ($RO_2$)

### Calculating background $RO_2$

@@

### Extracting data for $RO_2$

Typically, the analysis respirometry data requires the selection one or more linear sections of the raw data for more accurate calculations of $RO_2$ (e.g. Gordon et al., 1989, Dorey et al., 2014, Chu & Gal, 2017). The function `calc.rate()` can manually extract and process data segments from respiromety data. Data segments can be selected by (1) time period, (2) row numbers, (3) $O_2$ decrease or (4) proportion, which should accomodate most, if not all, data selection requirements and allow for consistent reporting of methods and results (e.g. estimating $RO_2$ over a regular 10-minute time window across multiple specimens). Rates are determined using linear regression, however a two-point analysis is also included alongside all regression results.

For more complex respirometry parameters, the function `auto.rate()` uses rolling regression and kernel density estimate algorithms to detect patterns in respirometry data. It can calculate maximum or minimum rates, extract rates at non-overlapping intervals, and has an exploratory method to detect the most linear sections of the data. When called, the function takes the data frame of length $m$, and performs a rolling regression of sample size $n$ iteratively across the length of the data (Fig. 2b). Thus, a total of $(m−n)+1$ number of overlapping regressions are fitted, and we obtain a new dataset of regressions as a function of time (rolling regression, Fig. 3b), which is then ranked by size to obtain maximum and minimum values.  There are clear advantages to using `auto.rate()` when estimating parameters for $MO_{2,max}$ or $MO_{2,min}$. Unlike manual methods, where data may be arbitrary selected, `auto.rate()` adheres to well-defined rules to obtain the rate of interest, so its methods are results are transparent and fully reproducible. To estimate the most linear sections of the data, further computations are performed after rolling regressions. Logically, linear sections of the data should reflect a consistently stable (i.e. flat) rate. The function takes advantage of this relationship and performs a Gaussian kernel density estimate of all regressions (see Silverman, 1986). New datasets around the kernel bandwidth are then resampled around peak values of the generated model, re-analysed, and then ranked by size to determine the most linear sections.

The function `calc.rate.ft()` is used to specifically analyse flowthrough respirometry data, and is similar to `calc.rate()` in functionality. However it accepts additional inputs of incurrent and excurrent oxygen concentration, and flow rate, that are not available in `calc.rate()`. Results in `calc.rate.ft()` are processed to determine the mean or modal $RO_2$.

## Estimating critical oxygen tension, $P_{crit}$

We provide two methods of estimating $P_{crit}$. The first is a "broken-stick" regression (BSR) approach, adopted from Yeager and Ultsch (1989) in which two segments of the data are iteratively fitted and the intersection with the smallest sum of the residual sum of squares between the two linear models is the estimated critical point. The second method is the segmented, or "broken-line" regression approach, presented by Muggeo (2003) which estimates the critical point by iteratively fitting two intersecting models on the data, and picking the point that minimises the "gap" between the two fitted lines. To determine $P_{crit}$, $RO_2$ (or $MO_2$) and ambient oxygen concentration ($PO_2$) must be provided. If $RO_2$ has not been calculated, `pcrit()` can automatically perform a rolling regression directly from raw respirometry data to generate the data and arrange it against $PO_2$, and then use one of the two methods to estimate the value (Fig. 4).

@@

For delibrate reasons, units of measurements are not used in any of the functions `calc.bg.rate()`, `calc.rate()`, `auto.rate()`, `calc.rate.ft()` and `pcrit()`. This is because
